Life in the canopy: community trait assessments reveal substantial functional diversity among fern epiphytes

### 2020-05-06 ms/manuscript.Rmd
Update MS to match corrections to proof

---
title: "Life in the canopy: Community trait assessments reveal substantial functional diversity among fern epiphytes"
bibliography: 
  - "references.bib"
  - "references_other.yaml"
mainfont: Roboto-Regular.ttf
fontsize: 11pt
mainfontoptions: 
- BoldFont=Roboto-Bold.ttf
- ItalicFont=Roboto-Italic.ttf
- BoldItalicFont=Roboto-BoldItalic.ttf
output: 
  bookdown::pdf_document2:
    citation_package: none # Use pandoc default, otherwise csl can't be used for reference formatting
    toc: no
    number_sections: no
    fig_caption: yes
    keep_tex: yes # need for converting tex to docx
    latex_engine: xelatex
    pandoc_args: [ "--csl", "new-phytologist.csl"]
    includes:
      in_header: new-phytologist.sty
header-includes: 
  - \usepackage{float}
  - \makeatletter\renewcommand*{\fps@figure}{H}\makeatother # Fix position of figs
  - \makeatletter\renewcommand*{\fps@table}{H}\makeatother # Fix position of tables
  - \usepackage{caption} # Needed to turn off automatic numbering of table captions
editor_options: 
  chunk_output_type: console
nocite: | 
  @Vogel1968
---

@@

^1^Department of Organismic and Evolutionary Biology and Harvard University Herbaria, Harvard University, Cambridge, MA 02138, U.S.A.; ^2^Department of Biology, Colgate University, 13 Oak Drive, Hamilton, New York 13346, U.S.A.; ^3^Current address: Department of Botany, National Museum of Natural History, Smithsonian Institution, Washington, D.C., 20013, U.S.A.

@@

<!-- FIXME: add borders to table in docx -->

@@

- While epiphytes showed convergence in small size and a higher frequency of non-cordate gametophytes, they showed greater functional diversity at the community level relative to terrestrial ferns.

@@

The appearance of angiosperm-dominated forests in the late Cretaceous and early Cenozoic fundamentally changed climates and had an immense effect on terrestrial ecosystems, including increased global precipitation and the creation of novel niches [@Niklas1983; @Kenrick1997a; @Lloyd2008; @Boyce2010]. 
The heterogeneous niche space of the angiosperm canopy comprising stratified light and humidity levels and a range of substrates for growth became rapidly filled by a variety of epiphytic plants, including mosses [@Fiz-Palacios2011], leafy liverworts [@Feldberg2014], orchids [@Givnish2015b], bromeliads [@Givnish2014a], and ferns [@Schneider2004d]. 
The transition to epiphytic growth was particularly important for modern fern species diversity: \circa{ }29% of all fern species are epiphytes vs. \circa{ }10% across all vascular plants [@Kress1986].
Moreover, major transitions to epiphytism occurred at least five times within ferns, each followed by diversification during the late Cretaceous or early Cenozoic [@Schuettpelz2009; @WatkinsJr.2012].
This has led to the hypothesis that epiphytism was a key factor contributing to extant fern diversification [@Schneider2004d; @Schuettpelz2009; @WatkinsJr.2012].
Recent studies exploring this idea [@Sundue2015; @Testo2016a; @Lehtonen2017; @Testo2018a], however, have largely failed to identify evidence of rapid diversification associated with epiphytism in ferns, calling into question the role of the canopy habitat in fern evolution.

While shifts in diversification rate are relevant, they are only one way epiphytic growth may have influenced fern evolution.
The impact of epiphytic growth could manifest in other ways, such as by shaping phylogenetic and functional diversity at the community level. 
For example, if the canopy is uniformly harsher than the terrestrial environment, we would expect functional convergence of epiphytes at the community level, which in turn may result in decreased phylogenetic diversity of epiphytic communities if traits are phylogenetically conserved [@Webb2002; @Cavender-Bares2009]. 
Conversely, the stratification of the canopy may stimulate functional diversification by niche partitioning, leading to greater functional diversity within epiphytic communities. 
Furthermore, the role of the canopy as a promoter or inhibitor of community diversity may be modulated by overall abiotic conditions, including most notably relative humidity (RH) and temperature. 
For instance, elevation is a strong predictor of fern species richness, with maximal richness occurring at mid-elevations, especially in the tropics [@Kromer2005; @Kessler2011a]. 
Functional diversity is also correlated with species richness [@Schleuter2010]. 
Thus, environmental variation along elevational gradients is expected to impact the functional, phylogenetic, and taxonomic diversity of epiphytic and terrestrial fern communities.

Fern epiphytes possess numerous adaptations that may contribute to their functional diversity.
In contrast to most terrestrial plants, epiphytes must contend with greater stress in the form of nutrient limitation, increased evaporative demand, drought, and extreme oscillations in temperature [@Zotz2001; @WatkinsJr.2009; @Lowman2012].
Epiphytic ferns tend to have morphological features that minimize water loss from their leaves, including shorter stipes [@WatkinsJr.2010], thicker leaves [@Kluge2007; @WatkinsJr.2007b], and protection with scales or hairs [@WatkinsJr.2006b]. 
Additional specialized adaptations to epiphytic growth include humus collection (\eg, *Asplenium*; @Benzing1990, *Drynaria*; @Janssen2005), vegetative desiccation tolerance (\eg, Hymenophyllaceae; @Shreve1911, *Pleopeltis*; @Pessin1925), CAM photosynthesis (\eg, *Pyrrosia*; @Wong1976), and symbiotic relationships with ants (\eg, *Antrophyum*; @WatkinsJr.2008, *Lecanopteris*; @Gay1993a).

While many similar adaptations are found in epiphytic seed plants and serve as key reference points [@Benzing1990], the evolution of epiphytism in ferns is more complex owing to their unique life-cycle. 
Ferns produce a free-living gametophyte that is typically smaller than the sporophyte and relies on aquatic, motile sperm.
Unlike the sporophyte, the fern gametophyte lacks vascular tissue, stomata, or a well developed cuticle, placing it in constant equilibrium with its external environment. 
In parallel with some of the adaptations of sporophytes, the gametophytes of some epiphytic ferns also exhibit morphological and physiological adaptations to minimize water loss and tolerate water scarcity. 
For example, extreme desiccation tolerance is prevalent in gametophytes of multiple lineages of epiphytic ferns, but is rare in sporophytes [@WatkinsJr.2007c; @Pittermann2013].
In addition, epiphytic fern gametophytes tend to exhibit complex three-dimensional morphologies that may function to retain water and slow its loss [@WatkinsJr.2007c].
Despite evidence for adaptations to epiphytism in multiple lineages in both sporophytes and gametophytes, no previous studies to our knowledge have analyzed the functional ecology of fern epiphytes in a broad phylogenetic context including both life stages.

Here, we seek to understand the role of the canopy in structuring fern community diversity by investigating putatively functional traits related to epiphytic growth in fern gametophytes and sporophytes on Moorea, French Polynesia.
Moorea is an isolated tropical oceanic island with high phylogenetic diversity of ferns distributed across a steep elevational gradient, making it ideal for a community phylogenetic analysis of functional traits. 
Specifically, we test the following hypotheses related to epiphytic growth in ferns: 1) Epiphytic ferns have convergent morphologies in both sporophyte and gametophyte phases; 2) Epiphytic fern communities have lower functional and phylogenetic diversity relative to terrestrial communities.

@@

Moorea, French Polynesia (17°32'S, 149°50'W) is a small (\SI{135}{\kilo\meter}^2^) tropical oceanic island more than \SI{5000}{\kilo\meter} from the nearest continental landmass.
This distance acts as a strong barrier to dispersal [@Carson1995; @Dassler2001], which combined with the young age of the island [\circa{ }\SI{1.5}{\ma}; @Duncan1976] has led to a relatively small, yet phylogenetically diverse fern flora [\circa{ }130 \species, 8/11 orders; `r s_table("species-list")`; taxonomy follows @PteridophytePhylogenyGroupI2016].
Ferns of Moorea include representatives of multiple Cretaceous epiphytic radiations [\eg, *Asplenium*, *Elaphoglossum*, Hymenophylloideae, Polypodiaceae, Vittarioideae; @Schuettpelz2009] and their terrestrial relatives, making them suitable for a comparative study of traits related to the evolution of epiphytic growth in ferns.

@@

A fern community survey was carried out on Moorea comprising 17, \SI{10}{\meter} × \SI{10}{\meter} plots spanning an elevational gradient from \SI{20}{\meter} to \SI{2000}{\meter}, with plots placed \circa every \SI{200}{\meter} [@Nitta2017].
Most plots were located along trails leading to the three main peaks on the island, Mt. Rotui (\SI{899}{\meter}), Mt. Mouaputa (\SI{890}{\meter}), and Mt. Tohiea (\SI{1206}{\meter}).
Each plot was divided into 25, \SI{2}{\meter} × \SI{2}{\meter} subplots.
Presence/absence of all sporophytes in each subplot was scored and summed to produce an abundance ranking for each species per plot from zero (not observed) to 25 (observed in all subplots).
Epiphytic ferns were sampled by hand to \circa{ }\SI{2}{\meter} on tree trunks or confirmed visually from the ground.
Species were categorized as epiphytic or terrestrial based on field observations (see 'Trait selection and measurements' below).
The list of species in each plot was split into epiphytic and terrestrial groups and treated separately during data analysis, hereafter referred to as 'epiphytic' and 'terrestrial' communities.
Vouchers were deposited at GH and UC [abbreviations follow Index Herbariorum; @Thiers2019].

@@

prelim_climate_survey <-
  # Subset to Moorea sites used in prelim survey (Mt. Tohiea only)
  climate_raw %>%
  filter(str_detect(site, "Aorai", negate = TRUE)) %>%
  filter(date < "2013-07-07") %>%
  filter(str_detect(site, "Tohiea")) %>%
  # Only count by site name, not distinguishing between epi vs. ter dataloggers
  mutate(site = str_remove_all(site, "_epi|_ter")) %>%
  summarize(
    n_sites = n_distinct(site),
    start = min(date) %>% 
      format('%B %e, %Y') %>%
      str_replace_all("  ", " "),
    end = max(date) %>% 
      format('%B %e, %Y') %>%
      str_replace_all("  ", " ")
  )

final_climate_survey <-
  # Subset to Moorea sites used in final survey
  climate_raw %>%
  filter(str_detect(site, "Aorai", negate = TRUE)) %>%
  filter(date >= "2013-07-07") %>%
  # Only count by site name, not distinguishing between epi vs. ter dataloggers
  mutate(site = str_remove_all(site, "_epi|_ter")) %>%
  summarize(
    n_sites = n_distinct(site),
    start = min(date) %>% 
      format('%B %e, %Y') %>%
      str_replace_all("  ", " "),
    end = max(date) %>% 
      format('%B %e, %Y') %>%
      str_replace_all("  ", " ")
  )

@@

Temperature and RH of epiphytic and terrestrial habitats were measured using Hobo Pro v2 dataloggers with the RS3 Solar Radiation Shield (Onset Corporation, USA).
A pair of dataloggers was installed for each plot, one mounted at \circa{ }\SI{2}{\meter} on a tree ('epiphytic datalogger'), and one at ground level ('terrestrial datalogger'), except for the Mt. Tohiea \SI{1200}{\meter} plot, which only had an epiphytic datalogger.
Temperature and RH were logged once every \SI{15}{\minute} during two survey periods: a preliminary survey from `r prelim_climate_survey$start` to `r prelim_climate_survey$end` at `r prelim_climate_survey$n_sites %>% english2()` plots, and a final survey from `r final_climate_survey$start` to `r final_climate_survey$end` at `r final_climate_survey$n_sites %>% english2()` plots.
Two plots were not included in the environmental survey.
We used the 2013--2014 data for the analysis, with one exception: the terrestrial datalogger at the Mt. Tohiea \SI{393}{\meter} plot malfunctioned and was missing data for a substantial part (March 12--July 5, 2014) of the final survey period, so we used data for the same dates from the preliminary survey period for this datalogger instead.
Three others also failed early during the final survey period, but lacked data from the preliminary period, so were excluded from analysis.
Other days missing data during the final survey period were excluded from the dataset.
The final dataset included `r total_days` days of data for `r total_sites %>% filter(habit == "epiphytic") %>% pull(n)` epiphytic dataloggers and `r total_sites %>% filter(habit == "terrestrial") %>% pull(n)` terrestrial dataloggers.
We calculated vapor pressure deficit (VPD) from relative humidity and temperature using the \package{plantecophys} R package [@Duursma2015].

@@

We selected several traits putatively relevant to epiphytic growth (`r table("traits-used")` and `r figure("traits-used-image")`).
We preferred morphological traits over physiological traits because data for morphological traits are easier to gather at the scale of our study (\circa{ }130 \species distributed over an elevational gradient from \SI{200}{\meter} to \SI{1200}{\meter}).
Plant morphology and physiology are closely linked, and physiological traits such as rates of photosynthesis and water use efficiency are known to be correlated with morphological traits such as stipe length in ferns [@WatkinsJr.2010].

Morphological traits of sporophytes were measured from herbarium specimens, or obtained from the literature when plant material was unavailable.
Measurements were made directly on plants in the field for a few species with large fronds.
A single leaf or rhizome per individual was measured on one to three separately collected individuals per species, and mean values were calculated for each species.
Although trait values can vary within fern species with elevation [@Kessler2007a], our study includes a broad phylogenetic sampling, so we assume intraspecific variation is outweighed by interspecies differences.
To measure specific leaf area (SLA; m^2^ kg^-1^), \circa{ }10--12, \SI{2}{\milli\meter} or \SI{4}{\milli\meter} diam. punches were taken of laminae between primary veins for a single leaf per individual using biopsy punches, dried at \SI{60}{\celsius} for \SI{24}{\hour}, and weighed.
For species with < \SI{2}{\milli\meter} between primary veins, leaf fragments excluding primary veins were obtained by dissection, dried at \SI{60}{\celsius} for \SI{24}{\hour}, weighed, and scanned.
ImageJ [@Abramoff2004] software was used to measure area per fragment.

Morphological traits of gametophytes were all qualitative (either categorical or binary; `r table("traits-used")`).
Traits were scored by observing gametophytes that were either collected in the field and identified using DNA barcoding [@Nitta2017], grown from spores of known species in the lab, taken from the literature, or inferred from taxonomic group (\ie, genus) if documented as a stable taxonomic character.
For comparative analysis, gametophyte morphotypes were aggregated into two categories: either cordate (\ie, heart-shaped) or non-cordate [@Ebihara2013].
Non-cordate morphotypes include elongate (strap or ribbon \sensu{ }@Farrar2008) and filamentous forms (`r figure("traits-used-image")`).
Gametophytes were cultivated on Bold's media [@Bold1957] supplemented with Nitch's micronutrients [@Nitsch1951].

Growth habit was coded as a binary trait (epiphytic or terrestrial) based on our field observations.
Some species do not fall clearly into either category [\eg, hemi-epiphytes; @Benzing1990; @Dubuisson2003; @Zotz2013].
We used connection to the soil as the criterion for defining binary growth habit; thus, epipetric plants were treated as epiphytic, and hemi-epiphytes and climbing plants were treated as terrestrial. 
For the purposes of this study, we treated growth habit as a fixed trait and used these categories to define epiphytic and terrestrial communities.
Other studies have demonstrated that exceptional epiphytic growth of terrestrial fern species (and vice-versa) is rare [@Cardelus2006; @Kluge2006; @WatkinsJr.2009], and we do not believe that such rare exceptions would affect our analysis.

@@

We calculated daily maximum, mean, minimum, and standard variation of temperature, RH, and VPD for each datalogger, then overall means from the daily values.
We analyzed correlation between the mean values with Pearson's correlation coefficient, and retained a set with correlation coefficients less than 0.9 (a single variable was arbitrarily selected from each set of correlated variables; hereafter, 'climatic variables').
Variables relating to VPD, which have no meaning if less than zero, and some other variables that had skewed distributions, were square-root transformed prior to fitting models.
We tested for differences between epiphytic and terrestrial dataloggers while accounting for elevation with ANCOVA, using growth habit as the categorical variable and elevation as the covariate.
We identified the best-fitting linear model for each climatic variable in response to growth habit and elevation by constructing a set of models including the effect of each independent variable alone, in combination, and in combination with their interactions. We then calculated the corrected Akaike Information Criterion [AIC~c~; @Akaike1973] for each model and retained models within 2 AIC~c~ of the lowest (\ie, 2 delta AIC~c~). From these, we selected the model with the combination of the fewest non-significant parameters and lowest AIC~c~ (for VPD, we the selected the model producing the smallest amount of negative predicted values).
A single outlier plot with unusually high temperature and low humidity (Mt. Rotui, 830 m plot) was excluded from ANCOVA and linear models.
Mt. Rotui is an isolated peak with greater exposure than the other two mountains where the rest of the plots were established (J. Nitta, pers. obs.).

For comparative analyses, we used the dated plastid phylogenetic tree of @Nitta2017.
*Microsorum maximum* (Brack.) Copel., a genetically verified hybrid between *M. grossum* (Langsd. & Fisch.) S. B. Andrews and *M. commutatum* Copel. [@Nitta2018] was excluded from analysis because it shares identical plastid sequences with *M. grossum*, so the plastid tree does not accurately represent its evolutionary distinctiveness.

>>>>

---
title: "Life in the canopy: community trait assessments reveal substantial functional diversity among fern epiphytes"
bibliography: 
  - "references.bib"
  - "references_other.yaml"
mainfont: Roboto-Regular.ttf
fontsize: 11pt
mainfontoptions: 
- BoldFont=Roboto-Bold.ttf
- ItalicFont=Roboto-Italic.ttf
- BoldItalicFont=Roboto-BoldItalic.ttf
output: 
  bookdown::pdf_document2:
    citation_package: none # Use pandoc default, otherwise csl can't be used for reference formatting
    toc: no
    number_sections: no
    fig_caption: yes
    keep_tex: yes # need for converting tex to docx
    latex_engine: xelatex
    pandoc_args: [ "--csl", "new-phytologist.csl"]
    includes:
      in_header: new-phytologist.sty
header-includes: 
  - \usepackage{float}
  - \makeatletter\renewcommand*{\fps@figure}{H}\makeatother # Fix position of figs
  - \makeatletter\renewcommand*{\fps@table}{H}\makeatother # Fix position of tables
  - \usepackage{caption} # Needed to turn off automatic numbering of table captions
editor_options: 
  chunk_output_type: console
nocite: | 
  @Vogel1968
---

@@

^1^Department of Organismic and Evolutionary Biology and Harvard University Herbaria, Harvard University, 22 Divinity Avenue, Cambridge, MA 02138, U.S.A.; ^2^Department of Biology, Colgate University, 13 Oak Drive, Hamilton, NY 13346, U.S.A.; ^3^Current address: Department of Botany, National Museum of Natural History, Smithsonian Institution, PO Box 37012, Washington, D.C. 20013, U.S.A.

@@

<!-- FIXME: add borders to table in docx -->
<!-- NOTE: the word count matches when the final, revised MS was submitted. It does not match changes made to the proof. -->

@@

- While epiphytes showed convergence in small size and a higher frequency of noncordate gametophytes, they showed greater functional diversity at the community level relative to terrestrial ferns.

@@

The appearance of angiosperm-dominated forests in the late Cretaceous and early Cenozoic fundamentally changed climates and had an immense effect on terrestrial ecosystems, including increased global precipitation and the creation of novel niches [@Niklas1983; @Kenrick1997a; @Lloyd2008; @Boyce2010]. 
The heterogeneous niche space of the angiosperm canopy comprising stratified light intensities and degrees of humidity and a range of substrates for growth became rapidly filled by a variety of epiphytic plants, including mosses [@Fiz-Palacios2011], leafy liverworts [@Feldberg2014], orchids [@Givnish2015b], bromeliads [@Givnish2014a], and ferns [@Schneider2004d]. 
The transition to epiphytic growth was particularly important for modern fern species diversity: \circa{ }29% of all fern species are epiphytes vs \circa{ }10% across all vascular plants [@Kress1986].
Moreover, major transitions to epiphytism occurred at least five times within ferns, each followed by diversification during the late Cretaceous or early Cenozoic [@Schuettpelz2009; @WatkinsJr.2012].
This has led to the hypothesis that epiphytism was a key factor contributing to extant fern diversification [@Schneider2004d; @Schuettpelz2009; @WatkinsJr.2012].
Recent studies exploring this idea [@Sundue2015; @Testo2016a; @Lehtonen2017; @Testo2018a], however, have largely failed to identify evidence of rapid diversification associated with epiphytism in ferns, calling into question the role of the canopy habitat in fern evolution.

While shifts in diversification rate are relevant, they are only one way epiphytic growth may have influenced fern evolution.
The impact of epiphytic growth could manifest in other ways, such as by shaping phylogenetic and functional diversity at the community level. 
For example, if the canopy is uniformly harsher than the terrestrial environment, we would expect functional convergence of epiphytes at the community level, which in turn may result in decreased phylogenetic diversity of epiphytic communities if traits are phylogenetically conserved [@Webb2002; @Cavender-Bares2009]. 
Conversely, the stratification of the canopy may stimulate functional diversification by niche partitioning, leading to greater functional diversity within epiphytic communities. 
Furthermore, the role of the canopy as a promoter or inhibitor of community diversity may be modulated by overall abiotic conditions, including relative humidity (RH) and temperature. 
For instance, elevation is a strong predictor of fern species richness, with maximal richness occurring at mid-elevations, especially in the tropics [@Kromer2005; @Kessler2011a]. 
Functional diversity is also correlated with species richness [@Schleuter2010]. 
Thus, environmental variation along elevational gradients is expected to impact the functional, phylogenetic, and taxonomic diversity of epiphytic and terrestrial fern communities.

Fern epiphytes possess numerous adaptations that may contribute to their functional diversity.
In contrast to most terrestrial plants, epiphytes must contend with greater stress in the form of nutrient limitation, increased evaporative demand, drought, and extreme oscillations in temperature [@Zotz2001; @WatkinsJr.2009; @Lowman2012].
Epiphytic ferns tend to have morphological features that minimize water loss from their leaves, including shorter stipes [@WatkinsJr.2010], thicker leaves [@Kluge2007; @WatkinsJr.2007b], and protection with scales or hairs [@WatkinsJr.2006b]. 
Additional specialized adaptations to epiphytic growth include humus collection (\eg{ }*Asplenium*; @Benzing1990, *Drynaria*; @Janssen2005), vegetative desiccation tolerance (\eg Hymenophyllaceae; @Shreve1911, *Pleopeltis*; @Pessin1925), Crassulacean acid metabolism (\eg{ }*Pyrrosia*; @Wong1976), and symbiotic relationships with ants (\eg{ }*Antrophyum*; @WatkinsJr.2008, *Lecanopteris*; @Gay1993a).

While many similar adaptations are found in epiphytic seed plants and serve as key reference points [@Benzing1990], the evolution of epiphytism in ferns is more complex owing to their unique life-cycle. 
Ferns produce a free-living gametophyte that is typically smaller than the sporophyte and relies on aquatic, motile sperm.
Unlike the sporophyte, the fern gametophyte lacks vascular tissue, stomata, or a well developed cuticle, placing it in constant equilibrium with its external environment. 
In parallel with some of the adaptations of sporophytes, the gametophytes of some epiphytic ferns also exhibit morphological and physiological adaptations to minimize water loss and tolerate water scarcity. 
For example, extreme desiccation tolerance is prevalent in gametophytes of multiple lineages of epiphytic ferns, but is rare in sporophytes [@WatkinsJr.2007c; @Pittermann2013].
In addition, epiphytic fern gametophytes tend to exhibit complex three-dimensional morphologies that may function to retain water and slow its loss [@WatkinsJr.2007c].
Despite evidence for adaptations to epiphytism in multiple lineages in both sporophytes and gametophytes, no previous studies have, to our knowledge, analyzed the functional ecology of fern epiphytes in a broad phylogenetic context including both life stages.

Here, we seek to understand the role of the canopy in structuring fern community diversity by investigating putatively functional traits related to epiphytic growth in fern gametophytes and sporophytes on Moorea, French Polynesia.
Moorea is an isolated tropical oceanic island with high phylogenetic diversity of ferns distributed across a steep elevational gradient, making it ideal for a community phylogenetic analysis of functional traits. 
Specifically, we test the following hypotheses related to epiphytic growth in ferns: 1) epiphytic ferns have convergent morphologies in both sporophyte and gametophyte phases; 2) epiphytic fern communities have lower functional and phylogenetic diversity relative to terrestrial communities.

@@

Moorea, French Polynesia (17°32'S, 149°50'W) is a small (\SI{135}{\kilo\meter}^2^) tropical oceanic island located more than \SI{5000}{\kilo\meter} from the nearest continental landmass.
This distance acts as a strong barrier to dispersal [@Carson1995; @Dassler2001], which combined with the young age of the island [\circa{ }\SI{1.5}{\ma}; @Duncan1976] has led to a relatively small, yet phylogenetically diverse fern flora [\circa{ }130 \species, 8/11 orders; `r s_table("species-list")`; taxonomy follows @PteridophytePhylogenyGroupI2016].
Ferns of Moorea include representatives of multiple Cretaceous epiphytic radiations [\eg{ }*Asplenium*, *Elaphoglossum*, Hymenophylloideae, Polypodiaceae, Vittarioideae; @Schuettpelz2009] and their terrestrial relatives, making them suitable for a comparative study of traits related to the evolution of epiphytic growth in ferns.

@@

A fern community survey was carried out on Moorea comprising 17 plots (\SI{10}{\meter} × \SI{10}{\meter}) spanning an elevational gradient from \circa{ }200 to \SI{1200}{\meter}, with plots placed \circa every \SI{200}{\meter} [@Nitta2017].
Most plots were located along trails leading to the three main peaks on the island, Mt. Rotui (\SI{899}{\meter}), Mt. Mouaputa (\SI{830}{\meter}), and Mt. Tohiea (\SI{1206}{\meter}).
Each plot was divided into 25 subplots (\SI{2}{\meter} × \SI{2}{\meter}).
Presence/absence of all sporophytes in each subplot was scored and summed to produce an abundance ranking for each species per plot from zero (not observed) to 25 (observed in all subplots).
Epiphytic ferns were sampled by hand to \circa{ }\SI{2}{\meter} on tree trunks or confirmed visually from the ground.
Species were categorized as epiphytic or terrestrial based on field observations (see 'Trait selection and measurements' later).
The list of species in each plot was split into epiphytic and terrestrial groups and treated separately during data analysis, hereafter referred to as 'epiphytic' and 'terrestrial' communities.
Vouchers were deposited at UC, with duplicates at GH and PAP [abbreviations follow Index Herbariorum; @Thiers2019].
All fieldwork was conducted under permits issued by the French Polynesian Government (Délégation à la Recherche) and the Haut-commissariat de la République en Polynésie Francaise (Protocole d'Accueil 2012–2014).

@@

prelim_climate_survey <-
  # Subset to Moorea sites used in prelim survey (Mt. Tohiea only)
  climate_raw %>%
  filter(str_detect(site, "Aorai", negate = TRUE)) %>%
  filter(date < "2013-07-07") %>%
  filter(str_detect(site, "Tohiea")) %>%
  # Only count by site name, not distinguishing between epi vs. ter dataloggers
  mutate(site = str_remove_all(site, "_epi|_ter")) %>%
  summarize(
    n_sites = n_distinct(site),
    start = min(date) %>% 
      format("%e %B %Y") %>%
      str_trim(),
    end = max(date) %>% 
      format("%e %B %Y") %>%
      str_trim()
  )

final_climate_survey <-
  # Subset to Moorea sites used in final survey
  climate_raw %>%
  filter(str_detect(site, "Aorai", negate = TRUE)) %>%
  filter(date >= "2013-07-07") %>%
  # Only count by site name, not distinguishing between epi vs. ter dataloggers
  mutate(site = str_remove_all(site, "_epi|_ter")) %>%
  summarize(
    n_sites = n_distinct(site),
    start = min(date) %>% 
      format("%e %B %Y") %>%
      str_trim(),
    end = max(date) %>% 
      format("%e %B %Y") %>%
      str_trim()
  )

@@

Temperature and RH of epiphytic and terrestrial habitats were measured using \textsc{Hobo Pro} v.2 dataloggers with the RS3 Solar Radiation Shield (Onset Corporation, Bourne, MA, USA).
A pair of dataloggers was installed for each plot, one mounted at \circa{ }\SI{2}{\meter} on a tree ('epiphytic datalogger'), and one at ground level ('terrestrial datalogger'), except for the Mt. Tohiea \SI{1200}{\meter} plot, which only had an epiphytic datalogger.
Temperature and RH were logged once every \SI{15}{\minute} during two survey periods: a preliminary survey from `r prelim_climate_survey$start` to `r prelim_climate_survey$end` at `r prelim_climate_survey$n_sites %>% english2()` plots, and a final survey from `r final_climate_survey$start` to `r final_climate_survey$end` at `r final_climate_survey$n_sites %>% english2()` plots.
Two plots were not included in the environmental survey.
We used the 2013--2014 data for the analysis, with one exception: the terrestrial datalogger at the Mt. Tohiea \SI{393}{\meter} plot malfunctioned and was missing data for a substantial part (12 March 2014 to 5 July 2014) of the final survey period, so we used data for the same dates from the preliminary survey period for this datalogger instead.
Three others also failed early during the final survey period, but lacked data from the preliminary period, so were excluded from analysis.
Other days missing data during the final survey period were excluded from the dataset.
The final dataset included \SI{`r total_days`}{\day} of data for `r total_sites %>% filter(habit == "epiphytic") %>% pull(n)` epiphytic dataloggers and `r total_sites %>% filter(habit == "terrestrial") %>% pull(n)` terrestrial dataloggers.
We calculated vapor pressure deficit (VPD) from RH and temperature using the \package{plantecophys} package in R [@Duursma2015].

@@

We selected several traits putatively relevant to epiphytic growth (`r table("traits-used")`; `r figure("traits-used-image")`).
We preferred morphological traits over physiological traits because data for morphological traits are easier to gather at the scale of our study (\circa{ }130 \species distributed over an elevational gradient from 200 to \SI{1200}{\meter}).
Plant morphology and physiology are closely linked, and physiological traits such as rates of photosynthesis and water use efficiency are known to be correlated with morphological traits such as stipe length in ferns [@WatkinsJr.2010].

Morphological traits of sporophytes were measured from herbarium specimens, or obtained from the literature when plant material was unavailable.
Measurements were made directly on plants in the field for a few species with large fronds.
A single leaf or rhizome per individual was measured on one to three separately collected individuals per species, and mean values were calculated for each species.
Although trait values can vary within fern species with elevation [@Kessler2007a], our study includes a broad phylogenetic sampling, so we assume intraspecific variation is outweighed by interspecies differences.
To measure specific leaf area (SLA; m^2^ kg^-1^), \circa{ }10--12, 2- or 4-mm-diameter punches were taken of laminae between primary veins for a single leaf per individual using biopsy punches, dried at \SI{60}{\celsius} for \SI{24}{\hour}, and weighed.
For species with < \SI{2}{\milli\meter} between primary veins, leaf fragments excluding primary veins were obtained by dissection, dried at \SI{60}{\celsius} for \SI{24}{\hour}, weighed, and scanned.
\package{ImageJ} [@Abramoff2004] software was used to measure area per fragment.

Morphological traits of gametophytes were all qualitative (either categorical or binary; `r table("traits-used")`).
Traits were scored by observing gametophytes that were either collected in the field and identified using DNA barcoding [@Nitta2017] or grown from spores of known species in the laboratory, taken from the literature, or inferred from taxonomic group (\ie genus) if documented as a stable taxonomic character.
For comparative analysis, gametophyte morphotypes were aggregated into two categories: either cordate (\ie heart-shaped) or noncordate [@Ebihara2013].
Noncordate morphotypes include elongate (strap or ribbon \sensu{ }@Farrar2008) and filamentous forms (`r figure("traits-used-image")`).
Gametophytes were cultivated on Bold's media [@Bold1957] supplemented with Nitch's micronutrients [@Nitsch1951].

Growth habit was coded as a binary trait (epiphytic or terrestrial) based on our field observations.
Some species do not fall clearly into either category [\eg hemi-epiphytes; @Benzing1990; @Dubuisson2003a; @Zotz2013].
We used connection to the soil as the criterion for defining binary growth habit; thus, epipetric plants were treated as epiphytic, and hemi-epiphytes and climbing plants were treated as terrestrial. 
For the purposes of this study, we treated growth habit as a fixed trait and used these categories to define epiphytic and terrestrial communities.
Other studies have demonstrated that exceptional epiphytic growth of terrestrial fern species (and vice-versa) is rare [@Cardelus2006; @Kluge2006; @WatkinsJr.2009], and we do not believe that such rare exceptions would affect our analysis.

@@

We calculated daily maximum, mean, minimum, and standard deviation (SD) of temperature, RH, and VPD for each datalogger, then overall means from the daily values.
We analyzed correlation between the mean values with Pearson's correlation coefficient, and retained those with correlation coefficients < 0.9 (a single variable was arbitrarily selected from each set of correlated variables; hereafter, 'climatic variables').
Variables relating to VPD, which have no meaning if less than zero, and some other variables that had skewed distributions, were square-root transformed prior to fitting models.
We tested for differences between epiphytic and terrestrial dataloggers while accounting for elevation with ANCOVA, using growth habit as the categorical variable and elevation as the covariate.
We identified the best-fitting linear model for each climatic variable in response to growth habit and elevation by constructing a set of models including the effect of each independent variable alone, in combination, and in combination with their interactions. 
We then calculated the corrected Akaike Information Criterion [AICc; @Akaike1973] for each model and retained models within 2 AICc of the lowest (\ie 2ΔAICc). 
From these, we selected the model with the combination of the fewest non-significant parameters and lowest AICc (for VPD, we selected the model producing the fewest negative predicted values).
A single outlier plot with unusually high temperature and low humidity (Mt. Rotui, 830 m plot) was excluded from ANCOVA and linear models.
Mt. Rotui is an isolated peak with greater exposure than the other two mountains where the rest of the plots were established (J. Nitta, pers. obs.).

For comparative analyses, we used the dated plastid phylogenetic tree of @Nitta2017.
*Microsorum* × *maximum* (Brack.) Copel., a genetically verified hybrid between *M. grossum* (Langsd. & Fisch.) S. B. Andrews and *M. commutatum* (Blume) Copel. was excluded from analysis because it shares identical plastid sequences with *M. grossum* [@Nitta2018], so the plastid tree does not accurately represent its evolutionary distinctiveness.

We tested for phylogenetic signal in quantitative traits using Blomberg's \K{ }[@Blomberg2003] and Pagel's \plam{ }[@Pagel1999], two metrics with slightly different interpretations and sensitivity to different evolutionary scenarios [@Munkemuller2012], with the \function{phylosig} function in the R package \package{phytools} [@Revell2012].
\plam is a scaling parameter that ranges from 0 (traits evolving randomly) to 1 (traits evolving along branches according to Brownian motion (BM)).
\K describes the ratio between the amount of observed variance in traits vs the amount of variance expected under BM: when \K{ }> 1, traits are more conserved than expected under BM; when \K{ }< 1, traits have less phylogenetic signal than expected under BM [@Blomberg2003].
We tested for significance in \K by comparing observed values of \K against a null distribution of \num{1000} trees with trait values randomized across the tips, and in \plam by conducting a likelihood ratio test comparing the log likelihoods of observed values of \plam vs \plam{ }= 0 (no phylogenetic signal).
We tested for phylogenetic signal in qualitative (binary) traits using Fritz and Purvis' \D{ }[@Fritz2010].
Values of \D range from 0 under trait evolution by BM to 1 under random distribution of traits, and can exceed these values in cases of extreme clumping (\D{ }< 0) or overdispersion (\D{ }> 1).
We analyzed \D using the \function{phylo.d} function in the R package \package{caper} [@Orme2013], and conducted a significance test by comparing observed \D to distributions of simulated values produced under two scenarios: random shuffling of traits on the tree, or simulation of a binary trait under a BM model.
To account for possible circularity in measuring the phylogenetic signal of traits that were scored based on taxonomy, we conducted the same analyses on a dataset excluding species that had been scored this way ('strict' dataset).
